Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2020-02-26 write-up/fda-recurrent.tex
Update fda-recurrent.tex

\begin{abstract}
Digital monitoring studies collect real-time high frequency data via mobile sensors in the subjects' natural environment. This data can be used to model the impact of changes in physiology on recurrent event outcomes such as smoking, drug use, alcohol use, or self-identified moments of suicide ideation. Likelihood calculations for the recurrent event analysis, however, become computationally prohibitive in this setting. Motivated by this, a random subsampling framework is proposed for computationally efficient, approximate likelihood-based estimation. A subsampling-unbiased estimator for the derivative of the cumulative hazard enters into an approximation of log-likelihood. The estimator has two sources of variation: the first due to the recurrent event model and the second due to subsampling. The latter can be reduced by increasing the sampling rate; however, this leads to increased computational costs.  The approximate score equations are equivalent to logistic regression score equations, allowing for standard, ``off-the-shelf'' software to be used in fitting these models. Simulations demonstrate the method and efficiency-computation trade-off. We then illustrate the approach using data from a digital monitoring study of suicidal ideation.
\end{abstract}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate choice of statistical model for joint high-frequency sensor data and the recurrent event data, and (2) to construct a simple, easy-to-implement method for parameter estimation and inference. For (1), we discuss an important issue regarding measurement-error models when paired with recurrent event outcomes. For (2), we introduce a random subsampling procedure that has several benefits.  First, the resulting inference is unbiased; however, there is a computation-efficiency trade-off. In particular, a higher sampling rate can decrease estimator variance at the cost of increased computation.  We show via simulations that the benefits of incredibly high sampling rates is often negligible, as the contribution to the variation is small in relation to to variation in the underlying stochastic processes. Second, derived estimating equations are optimal, implying loss of statistical efficiency is only due to subsampling procedure and not derived methodology.  Finally, implementation can leverage existing, standard software for functional data analysis and logistic regression, leading to fast adoption by domain scientists.

@@

Let~$H_{i,t}^{NX} = H_{i,t}^{N} \otimes H_{i,t}^{X}$ be the $\sigma$-field generated by all past values~$(N_i (s), x_i (s))_{0 \leq s \leq t}$. In this paper, the instantaneous risk of an event at time~$t$ is assumed to depend on the health process, time-in-study, and the event history through a fully parametric conditional hazard function:
\begin{equation}
\label{eq:hazard}
h_i \left( t \Given H_{i,t}^{NX} ; \theta \right) =
\lim_{\delta \to 0} \delta^{-1} \pr \left( N_i(t+\delta) - N_i(t) > 0
  \given H_{i,t}^{NX} \right),
\end{equation}
where~$\theta$ is the parameter vector. Equation~\eqref{eq:hazard} states the hazard may depend on past values of the event history and covariates.  For high frequency physiological data, we assume that current risk is log-additive and depends on a linear functional of the health process over some recent window of time and some pre-specified features of the counting process; that is,
\begin{equation}
\label{eq:hazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_i (s) \beta(s) ds  \right)
\end{equation}
where~$h_0(t;\gamma)$ is a parametrized baseline hazard function,~$\Delta$ is an unknown window-length, and $g_t( H_{i,t}^N ) \in \mathbb{R}^p$ is a $p$-length feature vector summarizing the event-history and time-in-study information. The final term~$\int_{t-\Delta}^t x_i(s) \beta(s) ds$ reflects the unknown linear functional form of the impact of the time-varying covariate on current risk.

@@

One potential criticism for~\eqref{eq:hazard} is that the health process may be measured with error. A common mathematical strategy for joint models is to consider an unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i \given \eta_i$, i.e., the two processes are conditionally independent given the latent trajectory. For example, take~$\eta_i$ to be a zero-mean Gaussian process with
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon_i(t)$ is a white-noise measurement error term. Thus, $\eta_i (t)$ is the ``true and unobserved value of the longitudinal outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional survival function of the $j$th event~$T_j$ given~$\bfx_i$ and all prior events~$\bfT_{-j} := \{ T_1, \ldots, T_{j-1} \}$ is
\[
\pr \left ( T_j > t+s \Given \H_{i,\tau_i}^X, \bfT_{-j} = \bft_{-j}, T_j > t \right) = E \left( - \exp \left( \int_{t}^{t+s} h_i (u \given \eta ) du \right) \Given \bfx_i, \bfT_{-j} = \bft_{-j} \right),
\]
where the expectation is a Gaussian integral, albeit infinite-dimensional.  For most choices of covariance structure, if the white-noise error term is non-zero then the above calculation will show the conditional survival function depends not only on past $x$-values, but also on future $x$-values, i.e.,~$\bfx$ and~$\bfT$ do not satisfy \emph{independent evolution}~\citep{DempseyPMCC2}. This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous risk of an event at time~$t$ depends on future values of the sensor process. To ensure independent evolution, we employ the strategy of setting~$\epsilon (t) \equiv 0$ and treating~$\bfx_i$ as a mean-zero Gaussian process. The proposed inferential procedure will only rely on the Gaussian assumption when~$\bfx$ is not fully observed (see Section~\ref{section:missingdata}).

@@

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. The component of the log-likelihood related to the event process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right)
\]
where~$h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

@@

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,t_i]$ denote the set of subsampled points.  Note, by construction, this set is distinct from the set of event times, i.e.,~$\bfT_i \cap \bfD_i = \emptyset$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (u) }
\]
An alternative design-unbiased estimator of the derivative of the cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Proposition~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting estimating equations can be written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in D_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} are approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

@@

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(s,t) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(s,t) = X(s,t) - \hat \mu_y(s,t)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (s,T), (s^\prime, T) ) f_y(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((s,t), (s^\prime,t))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$ and $f_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (s_r,t) \tilde X (s_{r^\prime}, t) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

Next, we take the spectral decomposition of the estimated covariance function; let $\{ \hat \psi^{(0)}_k (s),\hat \lambda^{(0)}_k \}_{k \geq 1}$ be the resulting sequence of eigenfunctions and eigenvalues. The key benefit of the marginal covariance approach is that it allows us to compute a single, time-invariant basis expansion; this reduces the computational burden by avoiding the three dimensional covariance function (i.e., covariance depends on~$t$) and associated spectral decomposition in methods considered by \cite{ChenMuller2012}. Using the Karhunen-Lo{\`e}ve decomposition, we can represent~$X(t,s)$ for $t \in \bfT_i \cap \bfD_i$ by
\[
X(s,t) = \hat \mu_y (s,t) + \sum_{k=1}^{\infty} \hat c^{(y)}_{i,k} (t) \hat
\psi^{(y)}_k (s) \approx \hat \mu_y (s,t) + {\bf c}^{(y)}_{i} (t)^\top \mathbold{\hat
  \psi}^{(y)} (s)
\]
where $\hat c^{(y)}_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat \psi^{(y)}_k (s) ds$, ${\bf c}^{(y)}_i (t) = (c^{(y)}_{i,1} (t), \ldots, c^{(y)}_{i,K_x} (t))^\top$, $\mathbold{\hat \psi}^{(y)} (s) = (\hat \psi^{(y)}_1 (s), \ldots, \hat \psi^{(y)}_{K_x} (s))^\top$, and~$K_x < \infty$ is the truncation level of the infinite expansion. Following~\cite{Goldsmith2011}, we set~$K_x$ to satisfy identifiability constraints (see Section~\ref{section:beta} for details). In subsequent sections, we leave the dependence on $y$ (i.e., whether $t \in \bfT_i$ or $\in \bfD_i$) implicit unless required for notational simplicity.

>>>>

\begin{abstract}
Digital monitoring studies collect real-time high frequency data via mobile sensors in the subjects' natural environment. This data can be used to model the impact of changes in physiology on recurrent event outcomes such as smoking, drug use, alcohol use, or self-identified moments of suicide ideation. Likelihood calculations for the recurrent event analysis, however, become computationally prohibitive in this setting. Motivated by this, a random subsampling framework is proposed for computationally efficient, approximate likelihood-based estimation. A subsampling-unbiased estimator for the derivative of the cumulative hazard enters into an approximation of log-likelihood. The estimator has two sources of variation: the first due to the recurrent event model and the second due to subsampling. The latter can be reduced by increasing the sampling rate; however, this leads to increased computational costs.  The approximate score equations are equivalent to logistic regression score equations, allowing for standard, ``off-the-shelf'' software to be used in fitting these models. Simulations demonstrate the method and efficiency-computation trade-off. We end by illustrating our approach using data from a digital monitoring study of suicidal ideation.
\end{abstract}

@@

The goal of this paper is two-fold: (1) to discuss the appropriate choice of statistical model for joint high-frequency sensor data and the recurrent event data, and (2) to construct a simple, easy-to-implement method for parameter estimation and inference. For (1), we discuss an important issue regarding measurement-error models when paired with recurrent event outcomes. For (2), we introduce a random subsampling procedure that has several benefits.  First, the resulting inference is unbiased; however, there is a computation-efficiency trade-off. In particular, a higher sampling rate can decrease estimator variance at the cost of increased computation.  We show via simulations that the benefits of incredibly high sampling rates is often negligible, as the contribution to the variation is small relative to the variation in the underlying stochastic processes. Second, derived estimating equations are optimal, implying loss of statistical efficiency is only due to subsampling procedure and not the derived methodology.  Finally, implementation can leverage existing, standard software for functional data analysis and logistic regression, leading to fast adoption by domain scientists.

@@

Let~$H_{i,t}^{NX} = H_{i,t}^{N} \otimes H_{i,t}^{X}$ be the $\sigma$-field generated by all past values~$(N_i (s), x_i (s))_{0 \leq s \leq t}$. In this paper, the instantaneous risk of an event at time~$t$ is assumed to depend on the health process, time-in-study, and the event history through a fully parametric conditional hazard function:
\begin{equation}
\label{eq:hazard}
h_i \left( t \Given H_{i,t}^{NX} ; \theta \right) =
\lim_{\delta \to 0} \delta^{-1} \pr \left( N_i(t+\delta) - N_i(t) > 0
  \given H_{i,t}^{NX} \right),
\end{equation}
where~$\theta$ is the parameter vector. For high frequency physiological data, we assume that current risk is log-additive and depends on a linear functional of the health process over some recent window of time and some pre-specified features of the counting process; that is,
\begin{equation}
\label{eq:hazardlinear}
h_i \left( t \given  H_{i,t}^{NX} ; \theta \right) =
h_0 (t; \gamma) \exp \left( g_t \left( H_{i,t}^{N} \right)^{\prime} \alpha
  + \int_{t-\Delta}^t x_i (s) \beta(s) ds  \right)
\end{equation}
where~$h_0(t;\gamma)$ is a parametrized baseline hazard function,~$\Delta$ is an unknown window-length, and $g_t( H_{i,t}^N ) \in \mathbb{R}^p$ is a $p$-length feature vector summarizing the event-history and time-in-study information. The final term~$\int_{t-\Delta}^t x_i(s) \beta(s) ds$ reflects the unknown linear functional form of the impact of the time-varying covariate on current risk.

@@

One potential criticism for~\eqref{eq:hazard} is that the health process may be measured with error. A common mathematical strategy for joint models is to consider an unobservable, latent process~$\eta_i$ such that~$\bfT_i \indep x_i \given \eta_i$, i.e., the two processes are conditionally independent given the latent trajectory. For example, take~$\eta_i$ to be a zero-mean Gaussian process with
\begin{equation}\label{eq:jm}
x_i(t) = \eta_i (t) + \epsilon_i (t),\quad \text{ and } \log h_i (t
\given \eta ) = \log h_0 (t) + g_t \left( H_{i,t}^N \right)^{\prime}
\alpha + \int_{t-\Delta}^t \eta_i (s) \beta (s) ds
\end{equation}
where~$\epsilon_i(t)$ is a white-noise measurement error term. Thus, $\eta_i (t)$ is the ``true and unobserved value of the longitudinal outcome'' \citep[Sec. 2.1, pp.3]{Rizopoulos2010}. The conditional survival function of the $j$th event~$T_j$ given~$\bfx_i$ and all prior events~$\bfT_{-j} := \{ T_1, \ldots, T_{j-1} \}$ is
\[
\pr \left ( T_j > t+s \Given \H_{i,\tau_i}^X, \bfT_{-j} = \bft_{-j}, T_j > t \right) = E \left( - \exp \left( \int_{t}^{t+s} h_i (u \given \eta ) du \right) \Given \bfx_i, \bfT_{-j} = \bft_{-j} \right),
\]
where the expectation is a Gaussian integral, albeit infinite-dimensional.  For most choices of covariance structure, if the white-noise error term is non-zero then the above calculation will show the conditional survival function depends not only on past $x$-values, but also on future $x$-values, i.e.,~$\bfx$ and~$\bfT$ do not satisfy \emph{independent evolution}~\citep{DempseyPMCC2}. This is quite unnatural, as~\eqref{eq:jm} suggests the instantaneous risk of an event at time~$t$ depends on future values of the sensor process. To ensure independent evolution, we employ the strategy of setting~$\epsilon (t) \equiv 0$.and treating~$\bfx_i$ as a Gaussian process measured without a white-noise error term. The proposed inferential procedure will only rely on the Gaussian assumption when~$\bfx$ is not fully observed (see Section~\ref{section:missingdata}).

@@

For the sake of notational simplicity, we leave the dependency of the conditional hazard function on~$H_{i,t}^{NX}$ implicit, and write~$h_i (t; \theta)$. The component of the log-likelihood related to the event process is then given by
\[
  L_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i}
    \log \left( h_i \left( T_{i,j}; \theta
      \right) \right) - H_{i} \left( \tau_i; \theta \right) \right)
\]
where~$H_{i} (\tau_i ; \theta) = \int_{0}^{\tau_i} h_{i} (t; \theta) dt$ is the cumulative hazard function. Solving the associated score equations~$U_n (\theta) = {\bf 0}$ yields the maximum likelihood estimator~$\hat \theta$, where
\[
U_n (\theta) = \sum_{i=1}^{n} \left ( \sum_{j=1}^{k_i} \frac{h^{(1)}_i
    (T_{i,j}; \theta)}{h_i (T_{i,j}; \theta)} - H^{(1)}_{i} (\tau_i;
  \theta) \right),
\]
with $h_i^{(1)} (T_{i,j}; \theta)$ and $H_i^{(1)} (\tau_{i}; \theta)$ are derivatives with respect to $\theta$.

@@

An estimator is design-unbiased if its expectation is equal to that parameter under the probability distribution induced by the sampling design~\citep{Cassel1977}. Let~$\bfD_i \subset [0,t_i]$ denote the set of subsampled points.  Note, by construction, this set is distinct from the set of event times, i.e.,~$\bfT_i \cap \bfD_i = \emptyset$.  Under subsampling via $\pi_i (t)$, one may compute a Horvitz-Thompson estimator of the derivative of the cumulative hazard:
\[
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in \bfD_i} \frac{
  h^{(1)}_i (u; \theta) }{ \pi_i (u) }
\]
An alternative design-unbiased estimator of the derivative of the cumulative hazards is given by
\begin{equation}
\label{eq:WPest}
\hat H_{i}^{(1)} (\tau_i; \theta) = \sum_{u \in (\bfT_i \cup \bfD_i)}
\frac{ h^{(1)}_i ( u; \theta ) }{ \pi_i (u) + h_i ( u ; \theta) }
\end{equation}
Equation~\eqref{eq:WPest} is the estimator suggested by~\cite{Waagepetersen2008}.  This estimator depends on the superposition of the event and subsampling processes. Proposition~\ref{prop:optimal} shows the estimator for~$\theta$ associated with using~\eqref{eq:WPest} is the most efficient within a suitable class of estimators for the derivative of the cumulative hazard function (including the Horvitz-Thompson estimator). Therefore, we restrict our attention to~\eqref{eq:WPest} for the remainder of this paper. Letting
\begin{equation}
\label{eq:waage_weights}
w_i (t; \theta) = \frac{\pi_i (t)}{\pi_i (t) + h_i (t ; \theta)},
\end{equation}
the resulting approximate estimating equations can be re-written as
\begin{equation}
\label{eq:approxscore}
\hat{U}_n (\theta) = \sum_{i=1}^n \left[ \sum_{u \in \bfT_i} w_i(u; \theta)
  \frac{h^{(1)} (u; \theta)}{ h ( u; \theta)}  - \sum_{u \in \bfD_i} w_i(u;
  \theta) \frac{h_i^{(1)} (u; \theta)}{ \pi_i (u) } \right].
\end{equation}
Equation~\eqref{eq:approxscore} represents the approximate score functions built via plug-in of the design-unbiased estimator of the derivative of the cumulative hazard given in~\eqref{eq:WPest}.

@@

We start by defining~$X(t,s) = x(t-s)$ to be the sensor measurement~$0 \leq s \leq \Delta$ time units prior to time~$t \in \bfT_i \cup \bfD_i$. We use the sandwich smoother~\citep{Xiao2013} to estimate the mean~$\mu_y(t,s) = \E_y [ X(t,s)]$ where the expectation is indexed by whether $t$ is an event ($y=1$) or subsampled ($y=0$) time respectively. Alternative bivariate smoothers exist, such as the kernel-based local linear smoother~\citep{Hastie2009} bivariate tensor product splines~\citep{Wood2006} and the bivariate penalized spline smoother~\citep{MarxEilers2005}. The sandwich smoother was chosen for its computational efficiency and estimation accuracy. We then define~$\tilde X(t,s) = X(t,s) - \hat \mu_y(t,s)$ to be the mean-zero process at each time~$t \in \bfT_i \cup \bfD_i$.

As in~\cite{Park2018}, define the \emph{marginal covariance} by
\[
\Sigma_y (s, s^\prime) = \int_{0}^\tau c_y( (T,s), (T,s^\prime) ) f_y(T) dt.
\]
for~$0 \leq s,s^\prime \leq \Delta$, where~$c_y((t,s), (t,s^\prime))$ is the covariance function of the windowed covariate history~$X(t,\cdot)$ and $f_y(T)$ is the intensity function for event ($y=1$) and subsampled ($y=0$) times respectively. Estimation of~$\Sigma_y$ occurs in two steps. For simplicity, we present the steps for subsampled times (i.e., $y=0$) but the steps are the same for event times as well. First, the pooled sample covariance is calculated at a set of grid points:
\[
\tilde \Sigma_0 (s_r , s_{r^\prime}) = \left( \sum_{i=1}^n \left | \bfD_i \right |\right)^{-1} \left( \sum_{i=1}^n \sum_{t \in \bfD_i} \tilde X (t,s_r) \tilde X (t,s_{r^\prime}) \right).
\]
Due to our concern over independent evolution as discussed in section~\ref{section:memproblems}, we do not assume that each observation is observed with white noise and therefore the diagonal elements of~$\hat \Sigma_0$ are not inflated. Second, the estimator~$\hat \Sigma$ is further smoothed again using the sandwich smoother~\citep{Xiao2013}. Note~\cite{Park2018} smooth the off-diagonal elements, while here we smooth the entire pooled sample covariance matrix. All negative eigenvalues are set to zero to ensure positive semi-definiteness. The result is used as an estimator~$\hat \Sigma_0$ for the pooled covariance $\Sigma_0$.

Next, we take the spectral decomposition of the estimated covariance function; let $\{ \hat \psi^{(0)}_k (s),\hat \lambda^{(0)}_k \}_{k \geq 1}$ be the resulting sequence of eigenfunctions and eigenvalues. The key benefit of the marginal covariance approach is that it allows us to compute a single, time-invariant basis expansion; this reduces the computational burden by avoiding the three dimensional covariance function (i.e., covariance depends on~$t$) and associated spectral decomposition in methods considered by \cite{ChenMuller2012}. Using the Karhunen-Lo{\`e}ve decomposition, we can represent~$X(t,s)$ for $t \in \bfT_i \cap \bfD_i$ by
\[
X(t,s) = \hat \mu_y (t,s) + \sum_{k=1}^{\infty} \hat c^{(y)}_{i,k} (t) \hat
\psi^{(y)}_k (s) \approx \hat \mu_y (t,s) + {\bf c}^{(y)}_{i} (t)^\top \mathbold{\hat
  \psi}^{(y)} (s)
\]
where $\hat c^{(y)}_{i,k} (t) = \int_{t-\Delta}^t \tilde X_i (s,t) \hat \psi^{(y)}_k (s) ds$, ${\bf c}^{(y)}_i (t) = (c^{(y)}_{i,1} (t), \ldots, c^{(y)}_{i,K_x} (t))^\top$, $\mathbold{\hat \psi}^{(y)} (s) = (\hat \psi^{(y)}_1 (s), \ldots, \hat \psi^{(y)}_{K_x} (s))^\top$, and~$K_x < \infty$ is the truncation level of the infinite expansion. Following~\cite{Goldsmith2011}, we set~$K_x$ to satisfy identifiability constraints (see Section~\ref{section:beta} for details). In subsequent sections, we leave the dependence on $y$ (i.e., whether $t \in \bfT_i$ or $\in \bfD_i$) implicit unless required for notational simplicity.
